# 4.5 Curves from lines

```{r}
library(rethinking)
data("Howell1")
d <- Howell1
```

The relationship in the whole population is not linear but curved.

```{r}
plot(height ~ weight, data=d)
```

## 4.5.1 Polynomial regression

Parabolic model of the mean

$\mu_i = \alpha + \beta_1x_i + \beta_2x_i^2$

$\beta_2$ measures the curvature of the relationship between weight and height.

We begin by **standardizing the predictor variable** - weight. - **Z-transform.** When predictor variables have very large values there are sometimes numerical glitches. **Especially important with polynomial regression where we square and cube the variable making it even larger**.

```{r}
d$weight_s <- (d$weight - mean(d$weight)) / sd(d$weight)
```

**The model**

$h_i \sim Normal(\mu_i, \sigma)$ [likelihood]

$\mu_i = \alpha + \beta_1x_i + \beta_2x_i$ [**polynomial regression model**]

$\alpha \sim Normal(178, 20)$ [$\alpha$ prior]

$\beta_1 \sim \text{Log-Normal}(0, 1)$ [$\beta_1$ prior]

$\beta_2 \sim Normal(0, 1)$ [$\beta_2$ **prior**]

$\sigma \sim Uniform(0, 50)$ [$\sigma$ prior]

We don't want to constraint the $\beta_2$ parameter to be positive like we did with $\beta_1$. Prior predictive simulation helps us understand why. In general, it is very difficult to understand the implication of the polynomial terms.

```{r}
d$weight_s2 <- d$weight_s^2
m4.5 <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b1*weight_s + b2*weight_s2 ,
        a ~ dnorm( 178 , 20 ) ,
        b1 ~ dlnorm( 0 , 1 ) ,
        b2 ~ dnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) , data=d )
```

```{r}
precis(m4.5)
```

Unlike the linear model the $\alpha$ here doesn't equal the sample mean. Why is that? Even when $x_i=0$ the average value of the quadratic term $E[\beta_2x_i^2]$ will likely not be equal to 0. Unlike with the linear model where $E[\beta_1x_i]$ was equal to 0.

**Linear model:**

```{r}
m4.5_linear <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b1*weight_s,
        a ~ dnorm( 178 , 20 ) ,
        b1 ~ dlnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) , data=d )

weight.seq <- seq( from=-2.2 , to=2 , length.out=30 )
pred_dat <- list( weight_s=weight.seq )
mu <- link( m4.5_linear , data=pred_dat )
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.89 )
sim.height <- sim( m4.5_linear , data=pred_dat )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )
```

```{r}
plot( height ~ weight_s , d , col=col.alpha(rangi2,0.5) )
lines( weight.seq , mu.mean )
shade( mu.PI , weight.seq )
shade( height.PI , weight.seq )
```

**Polynomial (quadratic) model**

```{r}
weight.seq <- seq( from=-2.2 , to=2 , length.out=30 )
pred_dat <- list( weight_s=weight.seq , weight_s2=weight.seq^2 )
mu <- link( m4.5 , data=pred_dat )
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.89 )
sim.height <- sim( m4.5 , data=pred_dat )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )
```

```{r}
plot( height ~ weight_s , d , col=col.alpha(rangi2,0.5) )
lines( weight.seq , mu.mean )
shade( mu.PI , weight.seq )
shade( height.PI , weight.seq )
```

**Polynomial (cubic) model**

```{r}
d$weight_s3 <- d$weight_s^3

m4.5_cubic <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + b1*weight_s + b2*weight_s2 + b3*weight_s3,
        a ~ dnorm( 178 , 20 ) ,
        b1 ~ dlnorm( 0 , 1 ) ,
        b2 ~ dnorm( 0 , 1 ) ,
        b3 ~ dnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) , data=d )

weight.seq <- seq( from=-2.2 , to=2 , length.out=30 )
pred_dat <- list( weight_s=weight.seq , weight_s2=weight.seq^2, weight_s3=weight.seq^3 )
mu <- link( m4.5_cubic , data=pred_dat )
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.89 )
sim.height <- sim( m4.5_cubic , data=pred_dat )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )

plot( height ~ weight_s , d , col=col.alpha(rangi2,0.5) )
lines( weight.seq , mu.mean )
shade( mu.PI , weight.seq )
shade( height.PI , weight.seq )
```

A better fit to the sample is could be a better geocentric (small world) model not guaranteed to be a good model in the big world (could be overfitting).

## 4.5.2 Splines

The second way to introduce curvature is to construct something known as a **spline.**

```{r}
data("cherry_blossoms")
d <- cherry_blossoms
precis(d)
```

[*doy*]{.underline} *here is the first day of blossom ranging from 86 (late March) to 124 (early May). The [year]{.underline} ranges frp, 812 CE to 2015 CE.*

```{r}
plot(doy ~ year, data=d)
```

**B-splines -** basis-splines. Basis is a component that is not wiggly. B-splines divide the full range of the predictor variable, like year, into parts. They assign a parameter to each part. These parameters are gradually turned on/off in a way that makes their sum into a fancy, wiggly curve.

Similar to polynomial regression, we generate new predictor variables and use them in the linear model for $\mu_i$. Unlike polynomial regression, B-splines do not directly transform the predictor by squaring or cubing it. Instead they invent a series of entirely new, synthetic predictor variables. Each exists only to gradually turn a specific parameter on/off within a specific range of the real predictor variable.

Each of these synthetic variables is called a **basis function**. The linear model becomes

$\mu_i = \alpha + w_1B_{i,1} + w_2B_{i,2} + w_3B_{i,3} + …$

where $B_{i, n}$ is the $n$-th basis function's value on row $i$, and $w$ parameters are the weights for each. These synthetic variable do some really elegant descriptive (geocentric) work for us.

How do we construct these basis functions B?

Divide the horizontal axis into parts, using pivot points called **knots**.

**Choose the knots**

Knots can be chosen in different ways. One approach is using the Wood's knot choice algorithm from the *mgcv* R package. We choose them here manually

```{r}
d2 <- d[complete.cases(d$doy), ]  # complete cases on doy
num_knots <- 15
knot_list <- quantile(d2$year, probs=seq(0, 1, length.out=num_knots))
```

**Choose polynomial degree**

For degree 1, two basis functions combine at each point. For degree 2, three basis functions combine at each point. For degree 3, four combine. etc.

```{r}
library(splines)
B <- bs(d2$year,
        knots=knot_list[-c(1, num_knots)],
        degree=3, intercept = T) # change degree=3 to something else to change the degree. Observe how many of the basis functions are non-zero
```

```{r}
plot( NULL , xlim=range(d2$year) , ylim=c(0,1) , xlab="year" , ylab="basis" )
for ( i in 1:ncol(B) ) lines( d2$year , B[,i] )
```

Notice how the knots become closer and closer. We chose the knots based on evenly-spaced quantiles. This gives more knots where there are more observations. It's very obvious that there are more observations in 1800-2000 than in 800-1200 and if we compare the pattern above we can see how concentrated the basis functions become the closer we are to the year 2000.

**Calculate parameter weights**

Now we need to get the parameter weights for each basis function. We need to define the model and fit it. The model is just a linear regression. The synthetic basis functions do all the work.

We also need an intercept. This would make it easier to define priors on the basis weights, because then we can just conceive of eacs as a deviation from the intercept.

$D_i \sim Normal(\mu_i, \sigma)$ [Likelihood]

$\mu_i = \alpha + \sum^K_{k=1}{w_kB_{k,i}}$ [linear model]

$\alpha \sim Normal(100, 10)$ [$\alpha$ prior]

$w_k \sim Normal(0, 10)$ [$w_k$ prior]

$\sigma \sim Exponential(1)$ [$\sigma$ prior]

```{r}
B
```
