# 6.3 Collider bias

Two factors play a role in the selection process for journal articles. They need to be trustowrthy and newsworthy. This is an example of **collider bias**.

T -\> S \<- N - this is an example of a collider - that is S.

When you condition on a collider it creates statistical, but not necessarily causal, associations among its causes. In this case, if we learn that a proposal has been selected (S), then learning its trustworthiness (T), also informs us about its newsworthiness (N).

Why?

Because if a proposal has been selected with low trustworthiness then its newsworthiness must be high because otherwise it wouldn't have been selected.

This information phenomenon therefore creates a negative association between T and N in the population of *selected* proposals.

**If you look at the selected proposals then low T would imply high N because otherwise it wouldn't be selected and vice versa. Also in the non-selected population high T would be associated with low N and vice versa because otherwise it would've been selected.**

## 6.3.1 Collider of false sorrow

How does aging influence happiness? If we survey people on how happy they are, is age associated with happiness? If so is it causal?

Suppose that happiness doesn't change as we age. Age does affect events in one's life such as marriage.

Happier people are more likely to get married: H-\>M

Older people are more likely to get married: A-\>M

So M is a collider: H-\>M\<-A

Happiness and age both cause marriage. So even if there is no causal association between age and happiness if we condition on M we will induce a statistical association. This can lead us to think that age influences happiness, when in fact it's constant.

```{r}
library(rethinking)
d <- sim_happiness(seed=1977, N_years = 1000)
precis(d)
```

Suppose you come across this data and want to find out whether age is related to happiness. You reason that marriage might be a confound because on average married people are more happy. So you include marriage status into the model to condition on it.

The linear model for this is

$\mu_i = \alpha_{MID[i]} + \beta_A A_i$

where $MID[i]$ is an index for the marriage status of individual i with 1 being single and 2 being married. It is a categorical variable coded with an index variable. It's easier to make priors when we use multiple intercepts, one for each category, than when we use indicator variables (0 and 1).

Now the priors:

Let's imagine that age and happiness have a very strong association and a person is happiest at 18 and declines until 65 when it's lowest. We rescale age so that 18 to 65 is one unit

```{r}
d2 <- d[d$age > 17,]
d2$A <- (d2$age - 18) / (65-18)
```

```{r}
library(patchwork)
library(tidyverse)

(d2 %>% ggplot(aes(x=d2$age)) + geom_density() + theme_classic()) + (d2 %>% ggplot(aes(x=d2$A)) + geom_density() + theme_classic())
```

Now A is the new variable. 0 means 18 years and 1 is 65 years.

Happiness is on a scale from -2 to +2.

```{r}
dens(d2$happiness)
```

```{r}
d2$mid <- d2$married + 1
m6.9 <- quap(
    alist(
        happiness ~ dnorm( mu , sigma ),
        mu <- a[mid] + bA*A,
        a[mid] ~ dnorm( 0 , 1 ),
        bA ~ dnorm( 0 , 2 ),
        sigma ~ dexp(1)
    ) , data=d2 )
precis(m6.9,depth=2)
```

Our model conditioning on marriage status shows that age is negatively associated with happiness.

If we now exclude marriage status.

```{r}
m6.10 <- quap(
    alist(
        happiness ~ dnorm( mu , sigma ),
        mu <- a + bA*A,
        a ~ dnorm(0, 1),
        bA ~ dnorm( 0 , 2 ),
        sigma ~ dexp(1)
    ) , data=d2 )
precis(m6.10,depth=2)
```

The model not conditioning on marriage status shows no association between age and happiness. This pattern is what we'd expect when we condition on a collider.

The collider is marriage status. It is a common consequence of age and happiness. Thus when we condition on it we induce a spurious association between the two causes.

## 6.3.2 The haunted DAG

Collider bias arises when conditioning on a common consequence. But sometimes it's hard to find a collider because there may be unmeasured causes. And that's something we should consider.
