# 6.4 Confronting confounding

**Confounding** is any context in which the association between an outcome Y and a predictor X is not the same as it would be if we experimentally determined X.

Suppose we're interested in the association between education E and wages W. In a population there are a lot of unsobserved variables U influencing both W and E.

```{r}
drawdag(dagitty("
                dag{
                U [unobserved]
                E -> W
                U -> E
                U -> W
                }
                "))
```

U can be where a person lives, their parents, their friends, etc.

If we regress W on E, its causal effect will be confounded by U because there are two paths connecting E and W:

-   the direct path W-\>E

-   the indirect (confounding) path E \<- W -\> U

We ignore the direction.

Only the first path is causal. Why? Because if only the second path existed and we changed E, it would not change W. DAG below:

```{r}
drawdag(dagitty("
                dag{
                U [unobserved]
                U -> E
                U -> W
                }
                "))
```

**How to isolate the causal path?**

We can run an experiment. If we assign values of E (we control it) the graph changes to the one below.

*U here doesn't affect E because we're the ones generating E. Therefore we're blocking the U-\>E path. And we've also blocked the E \<- U -\> W path (the non-causal path between E and W)*

```{r}
drawdag(dagitty("
                dag{
                U [unobserved]
                E -> W
                U -> W
                }
                "))
```

Now there's only one way for information to go between E and W and then measuring the association between E and W would give us a useful measure of the causal effect of E on W.

**Manipulation removed the confounding because it blocks the other path between E and W.**

There are statistical ways of achieving the same result without manipulating E.

The most obvious is to add U, to *condition* on it. It also blocks the path E \<- U -\> W. Why?

-   Once you learn U, also learning E doesn't give any additional information about W.

If U is the region a person lives in. Regions with high education and better paying jobs result in higher wages. If we don't know the region, then learning the individual's education tells us something about W because E and W are correlated across regions. But after we know where a person lives, learning E tells us nothing more about W. This is why conditioning on U blocks the E \<- U -\> W path. It makes E an W independent conditionally on U

## 6.4.1 Shutting the backdoor

Blocking confounding paths between predictor X and outcome Y is known as **shutting the backdoor.**

E \<- U -\> W is a backdoor path. U affecting E won't change W but it will cause a spurious association between E and W.

**Knowing the DAG it is [ALWAYS]{.underline} possible to say which variable we [SHOULD]{.underline} control for to shut all backdoor paths. It is also possible to say which variables we [SHOULDN'T]{.underline} control for.**

All we need to know are the **four elemental confounds.**

1.  **The fork** X\<-Z-\>Y.\
    Z is a common cause of both X and Y generating a correlation between them. If we condition on Z, learning X tells us nothing about Y. X and Y are independent conditional on Z
2.  **The pipe** X-\>Z-\>Y\
    Example is the post-treatment bias in plant growth. Treatment X influences fungus Z which influences growth Y. If we condition on Z now, we also block all paths from X to Y
3.  **The collider** X-\>Z\<-Y.\
    Unlike the others, there is **no association** between X and Y **unless** **you condition** on Z. Conditioning on Z opens that path. So we shouldn't include Z in the model.
4.  **The descendent**: The descendent is a fourth variable influenced by Z: Z-\> D. D carries some information from Z so conditioning on it partially opens the X-\>Z\<-Y path. It partially confounds the causal inference of X on Y. It depends on the nature of the relationship between Z and D Z-\>D.

The algorithm to follow to deal with confounds is:

1.  List all paths between X (potential cause of interest) and Y (the outcome)
2.  Classify each as open or closed: A path is open unless it contains a collider
3.  Classify each by whether it is a backdoor: A backdoor path has an arrow entering X.
4.  If there are any open backdoor paths decide on which ones to condition on to close it.

```{r}
dag_6.1 <- dagitty( "dag {
    U [unobserved]
    X -> Y
    X <- U <- A -> C -> Y
    U -> B <- C
}")

adjustmentSets( dag_6.1 , exposure="X" , outcome="Y" )
```
